---
title: 'Fig 1b: map of CKS by continent'
author: "Casey O'Hara"
date: "6/15/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(sf)
library(ggwaffle) ### devtools::install_github("liamgilbey/ggwaffle")
library(cowplot)
library(rnaturalearth)
```

# Summary

Generate a plot showing a continent map, each continent overlaid with a waffle plot showing the number and distribution of CKS by cultural group type.  

Is a map really necessary? because could also just do a facet grid.

# Methods

## Generate a continent panel

Very basic continent outline

```{r}
continents_sf <- rnaturalearth::ne_countries(returnclass = 'sf') %>%
  select(continent) %>%
  filter(continent != 'Antarctica')

cont_map <- ggplot() +
  geom_sf(data = continents_sf, aes(geometry = geometry),
          color = NA, fill = 'grey90') +
  theme_void()
```

## Generate waffle plots

Gather CKS counts by group type, then uncount them into individual observations - then `ggwaffle::waffle_iron()` to turn into a `geom_waffle`-ready dataframe

```{r}
cks_raw <- readxl::read_excel(here('data/CKS_Superclean.xlsx')) %>%
  janitor::clean_names() %>%
  filter(!is.na(continent)) %>%
  filter(cultural_group != 'Data deficient')

cult_gp_lbls <- cks_raw %>%
  group_by(cultural_group) %>%
  summarize(lbl = paste0(first(cultural_group), ' (n = ', n(), ')'))

cont_lbls <- cks_raw %>%
  group_by(continent) %>%
  summarize(lbl = paste0(first(continent), ' (n = ', n(), ')'))

cks_long_df <- cks_raw %>%
  mutate(continent_type = paste(continent, cultural_group, sep = 'X'))

cont_vec <- cks_long_df$continent %>% unique() %>% sort()
cks_waffle_df <- lapply(cont_vec,
    FUN = function(cont) {
      df <- cks_long_df %>%
        filter(continent == cont) %>%
        ggwaffle::waffle_iron(aes_d(group = continent_type), 
                              sample_size = 1) %>%
        separate(group, into = c('cont', 'cultural_group'), sep = 'X')
    }) %>%
  bind_rows() %>%
  left_join(cult_gp_lbls, by = 'cultural_group') %>%
  mutate(lbl = factor(lbl))

gen_plot <- function(df, continent) {
  brks <- levels(df$lbl)
  p <- ggplot() +
    geom_waffle(data = df %>% filter(cont == continent), 
                aes(y, -x, fill = lbl), size = .25, color = 'white') +
    scale_fill_manual(breaks = brks, values = hcl.colors(5)) +
    coord_equal() +
    theme_void() +
    theme(legend.key.size = unit(.35, 'cm')) +
    labs(fill = 'Each square represents\none cultural keystone species')
  return(p)
}

afr  <- gen_plot(df = cks_waffle_df, 'Africa')
s_am <- gen_plot(df = cks_waffle_df, 'South America')
n_am <- gen_plot(df = cks_waffle_df, 'North America')
asia <- gen_plot(df = cks_waffle_df, 'Asia')
eur  <- gen_plot(df = cks_waffle_df, 'Europe')
ocn  <- gen_plot(df = cks_waffle_df, 'Oceania')
glb  <- gen_plot(df = cks_waffle_df, 'Global')
```

## Combine plots

Parameterize x, y, width, height according to continent location and number of boxes
```{r}
afr_p <- get_panel(afr)
s_am_p <- get_panel(s_am)
n_am_p <- get_panel(n_am)
asia_p <- get_panel(asia)
eur_p <- get_panel(eur)
ocn_p <- get_panel(ocn)
glb_p <- get_panel(glb)
lgd <- get_legend(afr)

map_p <- get_panel(cont_map)
```

``` {r}
plot_w <- 10; plot_h <- 5; ar = plot_w / plot_h

hw_df <- cks_waffle_df %>%
  group_by(cont) %>%
  summarize(h = max(x), w = max(y)) %>%
  mutate(w = .08 * w/8,
         h = .08 * h/8 * ar,
         cont = tolower(cont) %>% str_sub(1, 3))

x <- list(afr = .5, ocn = .8, asia = .7, eur = .5, n_am = .2, s_am = .31)
y <- list(afr = .4, ocn = .19, asia = .6, eur = .67, n_am = .58, s_am = .25)
w <- as.list(hw_df$w) %>% setNames(hw_df$cont)
h <- as.list(hw_df$h) %>% setNames(hw_df$cont)

p <- ggdraw() +
  draw_plot(map_p, x = 0, y = 0, width = 1, height = 1) +
  draw_plot(afr_p, x = x$afr, y = y$afr, width = w$afr, height = h$afr) +
  draw_plot(eur_p, x = x$eur, y = y$eur, width = w$eur, height = h$eur) +
  draw_plot(asia_p, x = x$asia, y = y$asia, width = w$asi, height = h$asi) +
  draw_plot(n_am_p, x = x$n_am, y = y$n_am, width = w$nor, height = h$nor) +
  draw_plot(s_am_p, x = x$s_am, y = y$s_am, width = w$sou, height = h$sou) +
  draw_plot(ocn_p, x = x$ocn, y = y$ocn, width = w$oce, height = h$oce) +
  draw_plot(glb_p, x = .9, y = .6, width = w$glo, height = h$glo) +
  draw_label('Global\n(n = 4)', x = .903, y = .6, size = 11, hjust = 0, vjust = 1) +
  draw_label('North America\n(n = 89)', 
             x = x$n_am+.003, y = y$n_am, size = 11, hjust = 0, vjust = 1) +
  draw_label('South America\n(n = 68)', 
             x = x$s_am+.003, y = y$s_am, size = 11, hjust = 0, vjust = 1) +
  draw_label('Europe (n = 31)', 
             x = x$eur+.003, y = y$eur, size = 11, hjust = 0, vjust = 1) +
  draw_label('Africa (n = 63)', 
             x = x$afr+.003, y = y$afr, size = 11, hjust = 0, vjust = 1) +
  draw_label('Asia (n = 79)', 
             x = x$asia+.003, y = y$asia, size = 11, hjust = 0, vjust = 1) +
  draw_label('Oceania (n = 48)', 
             x = x$ocn+.003, y = y$ocn, size = 11, hjust = 0, vjust = 1) +
  draw_plot(lgd, x = .1, y = .05, width = .15, height = .15)

plot_f <- here('fig1b.png')
ggsave(plot_f, height = plot_h, width = plot_w, dpi = 300)

knitr::include_graphics(plot_f)
```

