---
title: 'Fig 1b: map of CKS by continent'
author: "Casey O'Hara"
date: "6/15/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(sf)
library(ggwaffle) ### remotes::install_github("liamgilbey/ggwaffle")
library(cowplot)
library(rnaturalearth)
```

# Summary

Generate a plot showing a continent map, each continent overlaid with a waffle plot showing the number and distribution of CKS by cultural group type.  

Also, add a panel A that shows waffle plot by taxonomic group.

# Methods: Panel A


```{r}
cks_raw <- readxl::read_excel(here('data/CKS_Superclean.xlsx')) %>%
  janitor::clean_names() %>%
  filter(!is.na(continent)) %>%
  filter(cultural_group != 'Data deficient') %>%
  group_by(cultural_group) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(cultural_group = fct_reorder(cultural_group, n)) %>%
  select(-n)
```

```{r}
tx_lbls <- cks_raw %>%
  select(taxa, species) %>%
  distinct() %>%
  group_by(taxa) %>%
  summarize(n = n()) %>%
  mutate(tx_lbl = sprintf('%s (n = %s)', taxa, n),
         tx_lbl = fct_reorder(tx_lbl, desc(n)))

nr <- 30 ### number of rows (actually columns...???) for panel A
tx_plot_df <- cks_raw %>%
  full_join(tx_lbls) %>%
  arrange(desc(tx_lbl))
  
nc <- ceiling(nrow(tx_plot_df)/nr) ### number of columns

grid_vals <- c(tx_plot_df$tx_lbl, 
               rep(NA, times = (nr * nc) - length(tx_plot_df$tx_lbl)))
grid_data <- expand.grid(y = 1:nr, 
                         x = 1:nc) 
grid_data <- grid_data %>%
  mutate(group = grid_vals) %>%
  filter(!is.na(group)) 

df <- grid_data %>%
  mutate(tx_lbl = levels(tx_plot_df$tx_lbl)[group],
         tx_lbl = factor(tx_lbl, levels = levels(tx_lbls$tx_lbl)))
```

``` {r make a discrete palette for 12 colors}
library(rcartocolor)

p12 <- carto_pal(12, "Safe") %>%
  colorspace::darken()
# swatch(p12)

```

``` {r}

panelA <- ggplot() +
  geom_waffle(data = df, 
              aes(y, -x, fill = rev(tx_lbl)), size = .25, color = 'white') +
  guides(fill = guide_legend(ncol = 2, reverse = FALSE) ) +
  scale_fill_manual(values = p12) +
  coord_equal() +
  theme_void() +
  theme(legend.key.size = unit(.35, 'cm'),
        legend.title = element_blank(),
        legend.position = 'right')

```


# Methods: Panel B

## Generate a continent panel

Very basic continent outline

```{r}
continents_sf <- rnaturalearth::ne_countries(returnclass = 'sf') %>%
  select(continent) %>%
  filter(continent != 'Antarctica')

cont_map <- ggplot() +
  geom_sf(data = continents_sf, aes(geometry = geometry),
          color = 'grey80', size = .25, fill = 'grey90') +
  theme_void()
```

## Generate waffle plots

Gather CKS counts by group type, then uncount them into individual observations - then `ggwaffle::waffle_iron()` to turn into a `geom_waffle`-ready dataframe

```{r redefine waffle_iron function}
waffle_iron2 <- function(df, rows = 8, na.rm = T) {

  df2 <- df %>%
    arrange(desc(cultural_group))
  cols <- ceiling(nrow(df2)/rows)
  grid_vals <- c(df2$continent_type, 
                 rep(NA, times = (rows * cols) - length(df2$continent_type)))
  grid_data <- expand.grid(y = 1:rows, 
                           x = 1:cols) 
  grid_data <- grid_data %>%
    mutate(group = grid_vals) %>%
    filter(!is.na(group))
  return(grid_data)
}
```

```{r}

cult_gp_lbls <- cks_raw %>%
  group_by(cultural_group) %>%
  summarize(lbl = paste0(first(cultural_group), ' (n = ', n(), ')')) %>%
  mutate(lbl = fct_reorder(lbl, desc(as.integer(cultural_group))))

cont_lbls <- cks_raw %>%
  group_by(continent) %>%
  summarize(lbl = paste0(first(continent), ' (n = ', n(), ')'))

cks_long_df <- cks_raw %>%
  mutate(continent_type = paste(continent, cultural_group, sep = 'X')) %>%
  arrange(cultural_group)

cont_vec <- cks_long_df$continent %>% unique() %>% sort()
cks_waffle_df <- lapply(cont_vec,
    FUN = function(cont) { ### cont = cont_vec[1]
      df <- cks_long_df %>%
        filter(continent == cont) %>%
        waffle_iron2() %>%
        separate(group, into = c('cont', 'cultural_group'), sep = 'X')
    }) %>%
  bind_rows() %>%
  left_join(cult_gp_lbls, by = 'cultural_group')

gen_plot <- function(df, continent) {
  brks <- levels(df$lbl)
  p <- ggplot() +
    geom_waffle(data = df %>% filter(cont == continent), 
                aes(y, -x, fill = lbl), size = .25, color = 'white') +
    scale_fill_manual(breaks = brks, values = hcl.colors(5)) +
    coord_equal() +
    theme_void() +
    theme(legend.key.size = unit(.35, 'cm')) +
    labs(fill = 'Each square represents\none culturally important species')
  return(p)
}

afr  <- gen_plot(df = cks_waffle_df, 'Africa')
s_am <- gen_plot(df = cks_waffle_df, 'South America')
n_am <- gen_plot(df = cks_waffle_df, 'North America')
asia <- gen_plot(df = cks_waffle_df, 'Asia')
eur  <- gen_plot(df = cks_waffle_df, 'Europe')
ocn  <- gen_plot(df = cks_waffle_df, 'Oceania')
glb  <- gen_plot(df = cks_waffle_df, 'Global')
```

## Combine plots

Parameterize x, y, width, height according to continent location and number of boxes
```{r}
afr_p <- get_panel(afr)
s_am_p <- get_panel(s_am)
n_am_p <- get_panel(n_am)
asia_p <- get_panel(asia)
eur_p <- get_panel(eur)
ocn_p <- get_panel(ocn)
glb_p <- get_panel(glb)
lgd <- get_legend(afr)

map_p <- get_panel(cont_map)
```

``` {r}

hw_df <- cks_waffle_df %>%
  group_by(cont) %>%
  summarize(h = max(x), w = max(y)) %>%
  mutate(w = .08 * w/8,
         h = .08 * h/8 * 2,
         cont = tolower(cont) %>% str_sub(1, 3))

x <- list(afr = .5, ocn = .8, asia = .7, eur = .5, n_am = .2, s_am = .31)
y <- list(afr = .4, ocn = .19, asia = .6, eur = .67, n_am = .58, s_am = .25)
w <- as.list(hw_df$w) %>% setNames(hw_df$cont)
h <- as.list(hw_df$h) %>% setNames(hw_df$cont)

panelB <- ggdraw() +
  draw_plot(map_p, x = 0, y = 0, width = 1, height = 1) +
  draw_plot(afr_p, x = x$afr, y = y$afr, width = w$afr, height = h$afr) +
  draw_plot(eur_p, x = x$eur, y = y$eur, width = w$eur, height = h$eur) +
  draw_plot(asia_p, x = x$asia, y = y$asia, width = w$asi, height = h$asi) +
  draw_plot(n_am_p, x = x$n_am, y = y$n_am, width = w$nor, height = h$nor) +
  draw_plot(s_am_p, x = x$s_am, y = y$s_am, width = w$sou, height = h$sou) +
  draw_plot(ocn_p, x = x$ocn, y = y$ocn, width = w$oce, height = h$oce) +
  draw_plot(glb_p, x = .87, y = .553, width = w$glo, height = h$glo) +
  draw_label('Multiple\ncontinents\n(n = 4)', x = .873, y = .55, size = 11, hjust = 0, vjust = 1) +
  draw_label('North America\n(n = 89)', 
             x = x$n_am+.003, y = y$n_am, size = 11, hjust = 0, vjust = 1) +
  draw_label('South America\n(n = 68)', 
             x = x$s_am+.003, y = y$s_am, size = 11, hjust = 0, vjust = 1) +
  draw_label('Europe (n = 31)', 
             x = x$eur+.003, y = y$eur, size = 11, hjust = 0, vjust = 1) +
  draw_label('Africa (n = 63)', 
             x = x$afr+.003, y = y$afr, size = 11, hjust = 0, vjust = 1) +
  draw_label('Asia (n = 79)', 
             x = x$asia+.003, y = y$asia, size = 11, hjust = 0, vjust = 1) +
  draw_label('Oceania (n = 48)', 
             x = x$ocn+.003, y = y$ocn, size = 11, hjust = 0, vjust = 1) +
  draw_plot(lgd, x = .55, y = .08, width = .15, height = .15)

```


```{r}

plot_f <- here('fig1_cis_spp_continent_top_btm.png')

waf_w <- 1
map_w <- 1
plot_w <- 8; plot_h <- 5

fig1 <- ggdraw() +
  draw_plot(panelA, x = 0, y = .73, width = waf_w, height = .25) +
  draw_plot(panelB, x = 0, y = 0, width = map_w, height = 0.75) +
  draw_label('A', x = 0.10, y = .95, hjust = 0, vjust = 1, size = 16, fontface = 'bold') +
  draw_label('B', x = 0.10, y = .64, hjust = 0, vjust = 1, size = 16, fontface = 'bold')

ggsave(plot_f, height = plot_h, width = plot_w, units = 'in', dpi = 600)

knitr::include_graphics(plot_f)
```

